---
title: Evaluating the Spread and Control of Zebra Mussels in U.S. Waterways
authors:
  - name: Brittany Bobb 
    affiliation: Colorado State 
    roles: writing
    corresponding: true
  - name: Louisa Beckett
    affiliation: Colorado State 
    roles: writing
    corresponding: true
  - name: Tayler Kordiak
    affiliation: Colorado State 
    roles: writing
    corresponding: true
  - name: Austin Sharrah
    affiliation: Colorado State 
    roles: writing
    corresponding: true
bibliography: references.bib
---

**Abstract**

Author : Louisa Beckett 

Zebra mussels (Dreissena polymorpha) pose a significant threat to freshwater ecosystems. This study investigates the impact of low-dose copper treatment (EarthTec QZ) on zebra mussel mortality in two separate bays within Lake Minnetonka, Minnesota: the relatively isolated St. Albans Bay and the more open Robinson Bay. We hypothesized that St. Albans Bay would exhibit higher copper concentrations and greater zebra mussel mortality due to reduced water mixing compared to Robinson Bay. Using a publicly available USGS dataset spanning July to October 2019, we analyzed trends in zebra mussel density and copper concentrations across five buoys in each bay. Faceted time-series plots revealed a substantial reduction in zebra mussel populations in St. Albans Bay following a 10-day copper treatment period, whereas Robinson Bay maintained significantly higher mussel densities. A paired t-test comparing copper levels at selected buoys in both bays demonstrated statistically significant differences (p = 0.0401), supporting the notion that the open morphology of Robinson Bay led to lower and more variable copper concentrations. While water quality parameters – dissolved oxygen, pH, and temperature – remained relatively stable, slight downward trends in dissolved oxygen and pH were observed in St. Albans Bay during treatment. Our findings suggest that bay morphology plays an important role in the effectiveness of localized copper treatments for zebra mussel control. It also highlights the importance of considering site-specific characteristics for targeted management strategies.



**Introduction**

Author: Tayler Kordiak

Zebra mussels (Dreissena polymorpha) are an aggressive bivalve invasive species that dominate and alter the freshwater ecosystems they invade. This benthic species, meaning living on the bottom of a body of water [@merriamwebster_benthic], often affects trophic levels by destroying ecosystem processes [@karatayev2002impacts]. Some ecosystem processes that are affected include oxygen and nutrient availability, sedimentation rates, and increased water temperature which can lead to a reduction of native species, due to outcompeting, that are detrimental to the freshwater ecosystems they belong to. Along with affecting the lake species, it also causes human problems economically and recreationally. Zebra mussels coined the term “biofouling” meaning that they attach to any hard surface such as rocks, boats, and docks because of their byssal threads [@nps_zebramussels_2025]. These mussels may not affect boats in the water for short trips, however, when these things are left in infested water for long periods, like power plant water intakes, it can cause millions of dollars in damage [@usgs_zebramussels_2025]. Zebra mussels were first brought into the United States in the mid-1980s, and it is hypothesized that they were brought from a cargo ship that ported in the St. Claire River, which connects Lake Erie and Lake Huron. The cargo ship, coming from Europe, dumped its ballast water which in turn dumped the organisms from the freshwater ports in Russia and Ukraine that they are native to. They were first discovered in Lake St. Claire, and quickly spread to the St. Croix and Mississippi Rivers naturally, and further to the west through human transport on boats [@nps_zebramussels_2025]. Currently, there are some preventative measures for halting the spread of infested waters to non-infested waters such as ensuring personal recreation equipment such as boats, jet skis, and trailers for these items are cleaned, drained, and well-dried before leaving the boat launch. During high-traffic months, there will typically be a USGS or State Department of Natural Resources monitor to inspect the boats that come in and out of the lakes to ensure that the proper safety measures are taken. This process includes inspecting equipment and removing any visible mussels, however, this may be difficult due to their small size, and draining live wells, bilges, and bait buckets of any water before departing. As an owner of this type of recreational equipment, it’s also your duty to ensure that your items are completely dry and washed before entering another body of water [@ca_water_mussels_2019]. 

Some current control methods are rapid response agencies, public education through signage and monitors at boat launches, as well as some physical, biological, and chemical controls. An example of a physical control measure is water drawdowns in canals and aqueducts which can dry and kill zebra mussels. There is ongoing biological research to determine the natural killers of zebra mussels, and the control measure that our group is interested in researching further is the chemical component that is being taken to kill off these mussels. The research question that we will be testing within the program R is, does the concentration of low-dose copper treatment in the treated St. Albans Bay in Lake Minnetonka in Minnesota affect zebra mussel mortality? We will be comparing this to an untreated Bay in the same lake, called Robinson Bay, and the data has been taken from the USGS website and initial low-dose copper treatment was applied in July of 2019, and ended in October of 2019. The work that we will be conducting is creating visual graphs for this experiment that was done in Lake Minnetonka. The objective was to maintain copper concentrations within St. Albans Bay to 60ug/L in the thermocline of the water for 10 days consecutively. Five testing sites were distinguished by buoys in each Bay. The goal was to target zebra mussel veligers (free-floating offspring) with the copper treatment called EarthTec QZ [@mnlra2019copper]. This is non-toxic for non-targeted animals within and outside of the lakes [@univar_earthtec_qz]. The data that will be used comes from the USGS Science-based catalog, which is a trusted website and displays public natural resource information for Lake Minnetonka, specifically, EarthTec QZ concentrations and zebra mussel density as well as mortality. We wanted to pick this lake because the testing zones will experience the same climate, topography, and prevention measures on the lake that are contaminated with zebra mussels. As for the methods, we will be using a log regression, decision tree, or random forest which will help us view the success of EarthTec QZ in St. Albans Bay compared to the control, Robinson Bay. We’ll want to apply regression analysis to view the statistical relationship between the treatments and the observed changes in the mussel population. Additionally, temporal trends will be assessed to evaluate the long-term impact of the control interventions. We will start with our full data, and pull out a sample of the data to be our training set. With the training set, we will be able to create resamples such as 10-fold cross-validation to compare the different models listed above to see which model best fits our data. Ensuring that each model has a specified model and engine. Once ranked, we will be able to decide which model we will use to show the density and mortality of zebra mussels within both lakes in Lake Minnetonka.



**Research Question**

How does copper treatment concentration affect zebra mussel mortality rates in bays with differing physical characteristics, such as the relative isolation of St. Albans Bay compared to the more open Robinson Bay?

**Hypothesis**

We hypothesize that St. Albans Bay will exhibit higher copper concentrations and lower zebra mussel populations than Robinson Bay due to its relative isolation from the rest of the lake, whereas the larger mouth of Robinson Bay will result in lower copper concentrations and will maintain higher mussel populations likely due to higher rates of water mixing.



**Methods**

Author: Brittany Bobb, Austin Sharrah

Using datasets from the USGS Science Data Catalog (SDC) publication on copper-based zebra mussel control in Lake Minnetonka, we aim to evaluate the effectiveness of a copper treatment by comparing zebra mussel mortality rates in Lake Minnetonka from before the time of treatment compared to after. This spatially referenced dataset enables us to explore how mussel mortality correlates with copper concentrations across treatment zones, offering insight into potential dose-dependent effects. The first bay, St. Alban’s Bay, is treated using a low-dose copper treatment, and will be compared with another nearby untreated bay, Robinson’s Bay. We mapped changes in zebra mussel distribution over a three month period July 2019 - October 2019, and applied regression analysis to examine the statistical relationship between treatment efforts and observed changes in mussel populations. The specific datasets we are utilizing are the Copper treatment data, Zebra Mussel Settlement data, and the Water Chemistry data. These datasets effectively cover the treatment type, specific bays, copper concentration, replication, time, date, and the water chemistry correlated to a specific data. While the data is largely structured, it did require cleaning such as handling duplicates, standardizing formatting, and addressing missing values. Verifying outliers was also addressed to ensure accurate analysis. To assess the effectiveness of copper treatment on zebra mussel control, we conducted a series of linear regression analyses using mussel settlement data collected from July to October 2019. Zebra mussel settlement was quantified by measuring the number of settled mussels per square meter (count/m²) at each buoy across both treated and untreated bays. The first regression model examined trends in mussel settlement over time, comparing settlement density (count/m²) with the timing and concentration of copper treatment applications (µg/L). This analysis enabled us to evaluate temporal patterns in response to treatment efforts, with copper application data overlaid to visualize the relationship between treatment events and observed changes in mussel populations. The model produced an R² value of [insert value], indicating the proportion of variation in mussel settlement that can be explained by the timing and concentration of copper applications. A second regression model assessed the relationship between mussel settlement and water chemistry parameters, including temperature and pH, alongside copper concentrations. This model was applied to data collected from individual buoys to account for spatial variability and to identify potential interactions between environmental conditions and treatment efficacy. The second model yielded an R² value of [insert value], reflecting the strength of association between environmental variables and settlement rates. Findings from these models suggest a positive correlation between copper concentration and zebra mussel mortality, with higher copper levels generally associated with reduced mussel settlement rates. These results provide evidence supporting the use of targeted copper treatments as an effective management strategy for zebra mussel populations, while also highlighting the influence of local environmental factors.



**Results**

Author: Tayler Kordiak

After making a faceted time series plot with Total (mussel) Count vs. Date (Figure 1) we can see that Robinson’s Bay (Figure 1, left) had significantly more zebra mussels than St. Albans Bay (Figure 1, right). However, the treatment in St. Albans Bay did a better job at almost eliminating the mussels completely from the Bay. Post-exposure, Robinson’s Buoy numbers 1, 3, and 5 all still had between 350,000 and 380,000 mussels, while St. Albans Buoys post-exposure had almost 0-200 mussels. Our hypothesis states, we believe St. Albans Bay copper treatment will be more effective because the bay’s opening to the rest of the lake is much smaller than that of Robinson’s Bay, thus assuming that St. Albans will have a lower zebra mussel distribution post-treatment. In that same note, we ran another faceted time series plot, Copper Levels vs. Date (Figure 2). We wanted to see whether or not the amount of the EarthTec QZ differed from bay to bay. As shown in Figure 2, there are obvious differences in the levels, Robinson’s Bay (Figure 2, left) had varying levels of copper surrounding each testing site between 2-23 μg/L with an outlier on August 1st which we can disregard, and St. Albans Bay  (Figure 2, right) had copper levels between 57-125 μg/L. These levels were pretty consistent during the 10 experimental days of July 21st and July 31st, but it is clear that St. Albans Bay had higher levels of the copper treatment, and our assumption from that is because of the openness of Robinson’s Bay compared to St. Albans Bay. The more mixing of untreated water with the treated water leads to a decrease in effectiveness. 


```{r}
library(dplyr)
mussel_settlement <- read.csv("C:/Users/bnb26/OneDrive/Desktop/ESS330/lab7/lab7/ZebraMusselSettlement.csv")
copper_data <- read.csv("C:/Users/bnb26/OneDrive/Desktop/ESS330/lab7/lab7/CopperData.csv")
```

```{r}
library(dplyr)
mussle_group <- mussel_settlement %>% mutate(count_numeric = as.numeric(Count.m.)) %>% group_by(Bay, Buoy.., Collection.Date) %>% summarize(total_count = sum(count_numeric, na.rm = TRUE), .groups = "drop") %>% select(Collection.Date, Buoy.., Bay, total_count)
```

```{r}
mussle_group <- mussle_group %>% rename(Buoy = Buoy..)
```

```{r}
library(ggplot2)

plot <- ggplot(mussle_group, aes(x = as.Date(Collection.Date, format = "%m/%d/%Y"), 
                        y = total_count, 
                        color = Buoy)) +
  geom_line(linewidth = 0.75) +
  facet_wrap(~ Bay, scales = "free_y") +
  labs(title = "Pre-exposure vs Post-exposure Mussel Count by Buoy",
       x = "Date", y = "Total Count") +
  theme_minimal() + theme(legend.position = "bottom")
print(plot)
```
*Figure 1. Faceted Time-Series Graphs of Robinson’s Bay (left) and St. Albans Bay (right) with Total Count on the y-axis and Date on the x-axis, split by Buoy*


```{r}
ggsave("C:/Users/bnb26/OneDrive/Desktop/ESS330/lab7/lab7/images/plot.png", plot = plot, width = 8, height = 6)
```

```{r}
copper_group <- copper_data %>% group_by(Bay, Buoy.., Copper..µg.L., Date) %>% summarise(copper_amount = sum(Copper..µg.L., na.rm = TRUE), .groups = "drop") %>% select(copper_amount, Buoy.., Bay, Date)
```

```{r}
copper_data <- copper_data %>% rename()
```

```{r}
copper_average <- copper_data %>%
  group_by(Bay, Buoy.., Date) %>%
  summarise(
    copper_amount = mean(Copper..µg.L., na.rm = TRUE),
    .groups = "drop"
  ) %>%
  select(copper_amount, Buoy.., Bay, Date)
```

```{r}
copper_average <- copper_average %>% rename(Buoy = Buoy..)
```

```{r}
copper_average <- copper_average %>% filter(Buoy != "Composite")
```

```{r}
copper_plot2 <- ggplot(copper_average, aes(x = as.Date(Date, format = "%m/%d/%Y"), 
                        y = copper_amount, 
                        color = Buoy)) +
  geom_line(linewidth = 0.35) +
  facet_wrap(~ Bay, scales = "free_y") +
  labs(title = "Copper Treatment Levels During Duration of Experiment",
       x = "Date", y = "Copper levels (µg/L)") +
  theme_minimal() + theme(legend.position = "bottom")
print(copper_plot2)
```

*Figure 2. Faceted Time-Series Graphs of Robinson’s Bay (left) and St. Albans Bay (right) with Copper levels (μg/L) on the y-axis and Date on the x-axis, split by Buoy*


```{r}
ggsave("C:/Users/bnb26/OneDrive/Desktop/ESS330/lab7/lab7/images/copper_plot2.png", plot = copper_plot2, width = 8, height = 6)
```

```{r}
library(tidyr)
library(dplyr)
paired_data4 <- copper_average %>%
  filter(Buoy %in% c("B4", "B5")) %>%
  pivot_wider(names_from = Buoy, values_from = copper_amount) %>%
  filter(!is.na(B4) & !is.na(B5))

t.test(paired_data4$B4, paired_data4$B5, paired = TRUE)

```


```{r}
library(readr)
library(tidyr)
library(dplyr)
water_chemistry <- read.csv("C:/Users/bnb26/OneDrive/Desktop/ESS330/lab7/lab7/WaterChemistry.csv")
```

```{r}
water_chemistry <- water_chemistry %>% rename(Temperature = Temperature...C.)
```

```{r}
library(tidyr)
library(ggplot2)
waterchem_long <- water_chemistry %>%
  pivot_longer(cols = c(pH, Dissolved.Oxygen..mg.L., Temperature),
               names_to = "variable",
               values_to = "value")

waterchem_long$variable <- recode(waterchem_long$variable,
                                  pH = "pH",
                                  Dissolved.Oxygen..mg.L. = "Dissolved Oxygen (mg/L)",
                                  Temperature = "Temperature (°C)")

supplemental_plots <- ggplot(waterchem_long, aes(x = as.Date(Date, format = "%m/%d/%Y"), y = value, color = Buoy..)) +
  geom_line() +
  facet_grid(rows = vars(variable), cols = vars(Bay), scales = "free_y") +
  labs(title = "Dissolved Oxygen, pH, and Temperature for Both Bays", x = "Date", y = NULL, color = "Buoy") +
  theme_minimal() +
  theme(strip.text = element_text(size = 6),
        legend.position = "bottom")
print(supplemental_plots)

```
*Figure 3. Faceted Time-Series Graphs showing Robinson’s Bay (left) St. Albans Bay (right) with Date on the x-axis and Dissolved Oxygen (mg/L) (top, left and right), pH (middle, left and right), and Temperature (℃) (bottom, left and right) on the y-axis*

```{r}
ggsave("C:/Users/bnb26/OneDrive/Desktop/ESS330/lab7/lab7/images/supplemental_maps.png", plot = supplemental_plots, width = 8, height = 6)
```
We realize that we have simply compared the two bays thus far using faceted time-series graphs, so we wanted to run a paired t-test, which could tell us if there is significance between the two bays. We decided to run the test on copper levels of Buoy 4 and Buoy 5 in both Robinson’s and St. Albans Bay to showcase statistical significance. After running the test, we got a p-value of 0.0401 and a t-value of 2.137 which is great and shows statistical significance (Table 1). There was clearly a difference between copper levels between bays, and more specifically between Buoy 4 and 5. Thus, we can reject the null hypothesis of there not being a difference in copper level between Robinson’s Bay and St. Albans Bay, and say with confidence that the openness to Lake Minnetonka of Robinson’s Bay can explain the lesser amount of copper at the five testing sites. Additionally, proving that the effectiveness of the copper treatment was more successful at the testing sites in St. Albans Bay. 
The copper treatment halted once the 10 days were up, this can be seen in Figure 2, and the levels of the copper started to decrease in both bays. This is a good thing because there is research out there that attests that extended use of EarthTec QZ can cause damage to the lake and aquatic life that reside there. This is because EarthTec QZ is copper in the form of biologically available copper, copper sulfate pentahydrate, which is acidic, thus turning the lake more acidic, or through killing algae, which will in turn consume the available oxygen within the lake [@showalter2019copper]. This is why we wanted to add supplemental data that was in the data we obtained from the original experiment, as exhibited by Figure 3. In Figure 3, again, we have another faceted time-series graph which contains the Robinson’s Bay (Figure 3, left) and St. Albans Bay (Figure 3, right) with graphs depicting Dissolved Oxygen (mg/L) (Figure 3, top left, top right), pH (Figure 3, middle left, middle right) and Temperature (℃) (Figure 3, bottom left, bottom right) for both bays. We felt it was necessary to add this supplemental data in our report because we wanted to see if the copper treatment would have an effect on the Dissolved Oxygen, pH, or Temperature of the water, or the other way around. We hypothesized that these three factors could have been confounding variables that may have affected the ability for the copper treatment to do its job. From Figure 3, we can see that their Dissolved Oxygen, pH, and Temperature were not affected and did not seem to make a difference on the copper level or the effectiveness of the treatment itself. However, in St. Albans Bay, Dissolved Oxygen (Figure 3, right, top) and pH (Figure 3, right, middle) did have a downward trend during the time of the experiment. As for Robinson’s Bay, both Dissolved Oxygen (Figure 3, left, top) and pH (Figure 3, left, middle) had pretty steady levels, averaging about the same throughout the experiment days. This can be explained by the higher levels of copper within St. Albans Bay from Figure 2. Temperature did not seem to be affected by the copper treatment or vice versa during the experiment. 




**Discussion & Conclusion**

Author: Lousia Beckett 

This study supports the hypothesis that bay morphology significantly influences the effectiveness of copper-based zebra mussel treatments. St. Albans Bay, which has a narrower connection to the rest of Lake Minnetonka, retained higher and more stable concentrations of EarthTec QZ, leading to a near-total reduction in zebra mussel populations across all five testing buoys. In contrast, Robinson’s Bay, with its wider mouth and greater rate of water exchange and mixing, maintained lower and more variable copper levels, resulting in substantially higher mussel densities post-treatment. This difference in treatment effectiveness is particularly visible after September 1st, when zebra mussel counts dropped sharply in St. Albans Bay whereas in Robinson’s Bay, the decrease was more gradual and had much less total change.
The statistical significance observed in the paired t-test between Buoys 4 and 5 in both bays (p = 0.0401) reinforces the conclusion that bay openness affects mussel mortality. The more open Robinson’s Bay likely experienced increased mixing with untreated water from the main lake, diluting copper concentrations and reducing treatment effectiveness. This finding directly supports our hypothesis and emphasizes how geographic and hydrologic context plays a vital role in the success of aquatic invasive species interventions.
Copper treatments don’t kill instantly; rather, they weaken zebra mussels over time, with mortality increasing as exposure lengthens. This is why copper application occurred in August while final population counts were recorded in September – to capture delayed treatment effects (Figures 1 & 2). The data also suggest that copper has a delayed effect, with zebra mussel mortality increasing over time as exposure accumulates. This relationship underscores the importance of maintaining effective treatment concentrations over a sustained period – in this case, the 10-day treatment window – to maximize impact on zebra mussel larvae and newly settled individuals.
While these findings are promising, there are limitations to our analysis. Both bays are connected to Lake Minnetonka, which introduces several potential confounding variables including water flow, organism dispersal, and external nutrients from runoff and such. Although St. Albans and Robinson’s Bays differ in their degree of isolation, both are influenced to some extent by lake-wide processes. This complicates our efforts to attribute population changes solely to treatment application. Environmental variables such as dissolved oxygen, pH, and temperature remained mostly stable during the study (Figure 3), but subtle downward trends in dissolved oxygen and pH in St. Albans Bay suggest potential interactions between copper exposure and aquatic chemistry that demand further investigation.
Despite these challenges, our results have broader relevance for environmental data science and invasive species management. They highlight the importance of controlled experimental design, careful site selection, and accounting for spatial variability when interpreting treatment outcomes. Replicating this study across different geographic settings and expanding the survey window beyond three months would help validate the effectiveness and ecological safety of EarthTec QZ in diverse aquatic systems.
This study provides evidence that low-dose copper treatments can be an effective, environmentally mindful method for zebra mussel control – especially when deployed in geographically favorable conditions such as partially closed off bays. These findings could inform predictive models used by lake managers to prioritize treatment sites based on bay geography and water exchange rates. Future work should continue to integrate spatial analysis, environmental monitoring, and advanced modeling techniques to optimize treatment strategies and reduce the ecological and economic impacts of invasive mussel populations.






